---
title: "Hands on Exercise 08c: Analytical Mapping"
author: "Yee Jun Yit"
date-modified: "last-modified"
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

## Overview

In this exercise, we will plot analytical maps.

We will learn how to:

-   Import an aspatial data file into R,
-   Convert it into simple point feature data frame and at the same time, to assign an appropriate projection reference to the newly create simple point feature data frame,
-   Plot interactive proportional symbol maps.

## Loading Data

We will be using the following main packages:

-   [**tmap**](https://r-tmap.github.io/tmap/) provides thematic maps, which are geographical maps in which spatial data distributions are visualized;
-   [**sf**](https://cran.r-project.org/web/packages/sf/sf.pdf) provides support for simple features, a standardized way to encode spatial vector data;

And as usual:

-   tidyverse, a family of R packages for data science processes.

```{r}
pacman::p_load(tmap, sf, tidyverse)
```

In this exercise, we will be using a prepared dataset *NGA_wp.rds*, which is in rds format. The data set is a polygon feature data.frame providing information on water points of Nigeria at the Local Government Area (LGA) level.

```{r}
NGA_wp <- read_rds("data/rds/NGA_wp.rds")
```

## Basic Choropleth Mapping

### Visualising distribution of non-functional water point

```{r}
#| fig-width: 12

p1 <- tm_shape(NGA_wp) +
  tm_fill(fill = "wp_functional",
          fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "matplotlib.blues"),
          fill.legend = tm_legend(position = tm_pos_out("right", "center"))) +
  tm_borders(lwd = 0.1,
             fill_alpha = 1) +
  tm_layout(main.title = "Distribution of functional water points by LGAs")

p2 <- tm_shape(NGA_wp) +
  tm_fill(fill = "total_wp",
          fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "matplotlib.blues"),
          fill.legend = tm_legend(position = tm_pos_out("right", "center"))) +
  tm_borders(lwd = 0.1,
             fill_alpha = 1) +
  tm_layout(main.title = "Distribution of total water points by LGAs")

tmap_arrange(p2, p1, nrow = 1)
```

### Choropleth Map for Rates

We have now seen the importance to map rates rather than counts of things, and that is for the simple reason that water points are not equally distributed in space. That means that if we do not account for how many water points are somewhere, we end up mapping total water point size rather than our topic of interest.

#### Deriving Proportion of Functional Water Points and Non-Functional Water Points

We will tabulate the proportion of functional water points and the proportion of non-functional water points in each LGA by using `mutate()` from **dplyr** package derive two fields: *pct_functional* and *pct_nonfunctional*.

```{r}
NGA_wp <- NGA_wp %>%
  mutate(pct_functional = wp_functional/total_wp) %>%
  mutate(pct_nonfunctional = wp_nonfunctional/total_wp)
```

### Plotting Map of Rates

We plot a choropleth map showing the distribution of percentage functional water points by LGA.

```{r}
#| fig-width: 12

tm_shape(NGA_wp) +
  tm_polygons(fill = "pct_functional",
          fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "matplotlib.blues"),
          fill.legend = tm_legend(title = "Percentage functional Water Pts"),
          fill.chart = tm_chart_histogram()) +
  tm_borders(fill_alpha = 1) +
  tm_layout(main.title = "Rate map of functional water point by LGAs")
```

## Extreme Value Maps

Extreme value maps are variations of common choropleth maps where the classification is designed to highlight extreme values at the lower and upper end of the scale, with the goal of identifying outliers. These maps were developed in the spirit of spatializing EDA, i.e. adding spatial features to commonly used approaches in non-spatial EDA (Anselin 1994).

### Percentile Map

A percentile map is a special type of quantile map with six specific categories: 0-1%,1-10%, 10-50%,50-90%,90-99%, and 99-100%. The corresponding breakpoints can be derived by means of the base R quantile command, passing an explicit vector of cumulative probabilities as c(0,.01,.1,.5,.9,.99,1). Note that the beginning and end points need to be included.

#### Data Preparation

Step 1: Exclude records with NA

```{r}
NGA_wp <- NGA_wp %>%
  drop_na()
```

Step 2: Create a custom classification and extract values

```{r}
percent <- c(0,.01,.1,.5,.9,.99,1)
var <- NGA_wp["pct_functional"] %>%
  st_set_geometry(NULL)
quantile(var[,1], percent)
```

::: {callout-warning}
##### Important

When variables are extracted from an sf data.frame, the geometry is extracted as well. For mapping and spatial manipulation, this is the expected behavior, but many base R functions cannot deal with the geometry. Specifically, the `quantile()` gives an error. Hence `st_set_geomtry(NULL)` is used to drop the geomtry field.
:::

#### Why Write Functions

Writing a function has three big advantages over using copy-and-paste:

-   We can give our function an evocative name that makes the code easier to understand.
-   As requirements change, we only need to update code in one place, instead of many places.
-   We eliminate the chance of making incidental mistakes which inevitably happens when we copy and paste (e.g. updating a variable name in one place, but not in another).

Source: (Wickham, R for Data Science)

#### Creating the get.var function

First, we will write an R function to extract a variable *wp_nonfunctional* as a vector out of a sf data.frame.

-   arguments:
    -   vname: variable name (as character, in quotes)
    -   df: name of sf data frame
-   returns:
    -   v: vector with values (without a column name)

```{r}
get.var <- function(vname,df) {
  v <- df[vname] %>% 
    st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}
```

#### A percentile mapping function

Next, we will write a percentile mapping function.

```{r}
percentmap <- function(vnam, df, legtitle=NA, mtitle="Percentile Map"){
  percent <- c(0,.01,.1,.5,.9,.99,1)
  var <- get.var(vnam, df)
  bperc <- quantile(var, percent)

  tm_shape(df) +
     tm_polygons(fill = vnam,
             fill.scale = tm_scale_intervals(
                breaks = bperc,
                values = "matplotlib.blues"),
             labels=c("< 1%", "1% - 10%", "10% - 50%", "50% - 90%", "90% - 99%", "> 99%"))  +
  tm_borders(fill_alpha = 1) +
  tm_layout(main.title = mtitle)
}
```

#### Test drive the percentile mapping function

We can now run the function

```{r}
percentmap("total_wp", NGA_wp)
```
Note that this is just a bare bones implementation. Additional arguments such as the title, legend positioning just to name a few of them, could be passed to customise various features of the map.

### Box map

A box map is in essence an augmented quartile map with an additional lower and upper category. When there are lower outliers, then the starting point for the breaks is the minimum value, and the second break is the lower fence. In contrast, when there are no lower outliers, then the starting point for the breaks will be the lower fence, and the second break is the minimum value i.e. there will be no observations that fall in the interval between the lower fence and the minimum value.

```{r}
ggplot(data = NGA_wp,
       aes(x = "",
           y = wp_nonfunctional)) +
  geom_boxplot()
```

We will be displaying summary statistics on a choropleth map by using the basic principles of boxplot.

To create a box map, a custom breaks specification will be used. However, there is a complication: the break points for the box map vary depending on whether lower or upper outliers are present.

#### Creating the boxbreaks function

We write an R function to create break points for a box map.

-  arguments:
   -   v: vector with observations
   -   mult: multiplier for IQR (default 1.5)
-  returns:
   -   bb: vector with 7 break points compute quartile and fences
   
```{r}
boxbreaks <- function(v,mult=1.5) {
  qv <- unname(quantile(v))
  iqr <- qv[4] - qv[2]
  upfence <- qv[4] + mult * iqr
  lofence <- qv[2] - mult * iqr
  # initialize break points vector
  bb <- vector(mode="numeric",length=7)
  # logic for lower and upper fences
  if (lofence < qv[1]) {  # no lower outliers
    bb[1] <- lofence
    bb[2] <- floor(qv[1])
  } else {
    bb[2] <- lofence
    bb[1] <- qv[1]
  }
  if (upfence > qv[5]) { # no upper outliers
    bb[7] <- upfence
    bb[6] <- ceiling(qv[5])
  } else {
    bb[6] <- upfence
    bb[7] <- qv[5]
  }
  bb[3:5] <- qv[2:4]
  return(bb)
}
```

#### Creating the get.var function

We now write a R function to extract a variable as a vector out of an sf data frame.

-  arguments:
   -   vname: variable name (as character, in quotes)
   -   df: name of sf data frame
-  returns:
   -   v: vector with values (without a column name)
   
```{r}
get.var <- function(vname,df) {
  v <- df[vname] %>% st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}
```

#### Test drive the newly created function

```{r}
var <- get.var("wp_nonfunctional", NGA_wp) 
boxbreaks(var)
```

#### Boxmap function

Below is an R function to create a box map.

-  arguments:
   -   vnam: variable name (as character, in quotes)
   -   df: simple features polygon layer
   -   legtitle: legend title
   -   mtitle: map title
   -   mult: multiplier for IQR 
-  returns: 
   -   a tmap-element (plots a map)

```{r}
boxmap <- function(vnam, df, 
                   legtitle=NA,
                   mtitle="Box Map",
                   mult=1.5){
  var <- get.var(vnam,df)
  bb <- boxbreaks(var)
  tm_shape(df) +
    tm_polygons() +
  tm_shape(df) +
     tm_polygons(fill = vnam,
                 fill.scale = tm_scale_intervals(
                   breaks = bb,
                   values = "matplotlib.blues"),
                 labels = c("lower outlier", 
                            "< 25%", 
                            "25% - 50%", 
                            "50% - 75%",
                            "> 75%", 
                            "upper outlier"))  +
  tm_borders() +
  tm_layout(main.title = mtitle)
}
```

```{r}
tmap_mode("plot")
boxmap("wp_nonfunctional", NGA_wp)
```


## References

-   [tmap](https://cran.r-project.org/web/packages/tmap/tmap.pdf) package documentation
-   [Chapter 25: Functions](https://r4ds.hadley.nz/functions.html), **R for Data Science** (Wickham)
