---
title: "Hands On Ex 00: Working with Tidyverse"
author: "Yee Jun Yit"
date-modified: "last-modified"
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

## Getting started
Load tidyverse into r env by using code chunk below
freeze parameter prevents running of code chunks 

```{r}
pacman::p_load(tidyverse)
```

Importing data (Old way - deprecated)
```{r}
#| eval: FALSE
realis_csv <- read.csv("data/REALIS2019.csv")
```

## Importing data (New way)
Uses readr to import csv as a tibble data.frame

```{r}
realis2019 <- read_csv("data/REALIS2019.csv")
```
```{r}
popdata_fat <- read_csv("data/PopData2019_fat.csv")
```
## Pivoting data
%>% sequentially processes functions, called "pipe"
```{r}
popdata_long = popdata_fat %>%
  pivot_longer(c(3:21),
               names_to = "Age Group",
               values_to = "Population")
```

## Writing and reading the file
rds is a r data serialisation
*Remember to add rds folder to .gitignore as it is a data folder

```{r}
write_rds(popdata_long, "rds/popdata_long.rds")
```

```{r}
popdata_long = read_rds("rds/popdata_long.rds")
```

## Working with dplyr package

### select() function
```{r}
realis2019_selected = realis2019 %>%
  select(`Project Name`,
         `Type of Sale`,
         `Unit Price ($ psm)`,
         `Property Type`)
realis2019_selected
```
### filter() function
```{r}
realis2019_filtered = realis2019_selected %>%
  filter(`Property Type` =="Condominium" |
          `Property Type`== "Apartment") %>%
  filter(`Type of Sale` == "New Sale") %>%
  filter(`Unit Price ($ psm)` <= 13000)
```

### combining both functions
```{r}
realis2019_combined = realis2019 %>%
  select(`Project Name`,
         `Type of Sale`,
         `Unit Price ($ psm)`,
         `Property Type`) %>%
  filter(`Property Type` =="Condominium" |
          `Property Type`== "Apartment") %>%
  filter(`Type of Sale` == "New Sale") %>%
  filter(`Unit Price ($ psm)` <= 13000)
realis2019_filtered
```
------------------------------------------------------------------------
