---
title: "Hands On Ex 00: Working with Tidyverse"
author: "Yee Jun Yit"
date-modified: "last-modified"
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

# Getting started:

## Load tidyverse into r env by using code chunk below

```{r}
pacman::p_load(tidyverse)
```

Importing data (Old way - deprecated)

```{r}
#| eval: FALSE
realis_csv <- read.csv("data/REALIS2019.csv")
```

## Importing data (New way) - uses readr to import as a tibble data.frame

```{r}
realis2019 <- read_csv("data/REALIS2019.csv")
```
```{r}
popdata_fat <- read_csv("data/PopData2019_fat.csv")
```
## Pivoting data (the indented function is processed on popdata_fat, called "pipe")
```{r}
popdata_long = popdata_fat %>%
  pivot_longer(c(3:21),
               names_to = "Age Group",
               values_to = "Population")
```

## Writing the file to a r data serialisation, remember to add rds folder to .gitignore as it is a data folder

```{r}
write_rds(popdata_long, "rds/popdata_long.rds")
```

```{r}
popdata_long = read_rds("rds/popdata_long.rds")
```

## Working with dplyr:

### select function
```{r}
realis2019_selected = realis2019 %>%
  select('Project Name',
         'Type of Sale',
         'Unit Price ($ psm)',
         'Property Type')
realis2019_selected
```
### filter function
```{r}
realis2019_filtered = realis2019_selected %>%
  filter('Property Type' =="Condominium" |
          'Property Type'== "Apartment") %>%
  filter('Type of Sale' == "New Sale") %>%
  filter('Unit Price ($ psm)' <= 13000)
```
###combining both functions
```{r}
realis2019_combined = realis2019 %>%
  select('Project Name',
         'Type of Sale',
         'Unit Price ($ psm)',
         'Property Type') %>%
  filter('Property Type' =="Condominium" |
          'Property Type'== "Apartment") %>%
  filter('Type of Sale' == "New Sale") %>%
  filter('Unit Price ($ psm)' <= 13000)
realis2019_combined
```
------------------------------------------------------------------------
