---
title: "Hands on Exercise 10: Information Dashboard Design: R methods"
author: "Yee Jun Yit"
date-modified: "last-modified"
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

## Overview

In this exercise, we will:

-   create bullet charts by using **ggplot2**,
-   create sparklines by using **ggplot2**,
-   build dashboards according to industry standards by using R Shiny.

## Loading Data

We will be using the following network data modelling and visualisation packages:

-   [**lubridate**](https://lubridate.tidyverse.org/) part of **tidyverse**, makes it easier to do the things R does with date-times and possible to do the things R does not;
-   [**ggthemes**](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/), an extension of **ggplot2**, provides additional themes beyond the basic themes of ggplot2.;
-   [**gtExtras**](https://jthomasmock.github.io/gtExtras/index.html/) provides some additional helper functions to assist in creating beautiful tables with [**gt**](https://gt.rstudio.com/), an R package specially designed for anyone to make wonderful-looking tables using R ;
-   [**reactable**](https://glin.github.io/reactable/index.html/) provides functions to create interactive data tables for R, based on [React Table](https://github.com/tanstack/table/tree/v7) and made with [reactR](https://github.com/react-R/reactR);
-   [**reactablefmtr**](https://kcuilla.github.io/reactablefmtr/index.html) provides various features to streamline and enhance the styling of interactive reactable tables with easy-to-use and highly-customizable functions and themes;

And as usual:

-   tidyverse, a family of R packages for data science processes.

```{r}
pacman::p_load(lubridate, ggthemes, gt, gtExtras, svglite, reactable, reactablefmtr,
               tidyverse)
```

In this exercise, we will be using a personal database in Microsoft Access mdb format called *Coffee Chain*.

We will use `odbcConnectAccess()` of [RODBC](https://cran.r-project.org/web/packages/RODBC/index.html) package to import a database query table into R.

```{r}
library(RODBC)
con <- odbcConnectAccess2007('data/Coffee Chain.mdb')
coffeechain <- sqlFetch(con, 'CoffeeChain Query')
write_rds(coffeechain, "data/CoffeeChain.rds")
odbcClose(con)
```

::: {callout-note}
### Note

Before running the code chunk, we need to change the R system to 32bit version in Tools -\> Global Options. This is because the `odbcConnectAccess()` is based on 32bit and not 64bit R.
:::

We can now make use of *CoffeeChain.rds*

```{r}
coffeechain <- read_rds("data/CoffeeChain.rds")
```

We will aggregate Sales and Budgeted Sales at the Product level for later use.

```{r}
product <- coffeechain %>%
  group_by(`Product`) %>%
  summarise(`target` = sum(`Budget Sales`),
            `current` = sum(`Sales`)) %>%
  ungroup()
```

## Bullet chart in ggplot2

The below plots bullet charts using ggplot2 functions.

```{r}
ggplot(product, aes(Product, current)) + 
  geom_col(aes(Product, max(target) * 1.01),
           fill="grey85", width=0.85) +
  geom_col(aes(Product, target * 0.75),
           fill="grey60", width=0.85) +
  geom_col(aes(Product, target * 0.5),
           fill="grey50", width=0.85) +
  geom_col(aes(Product, current), 
           width=0.35,
           fill = "black") + 
  geom_errorbar(aes(y = target,
                    x = Product, 
                    ymin = target,
                    ymax= target), 
                width = .4,
                colour = "red",
                size = 1) +
  coord_flip()
```

## Sparklines using ggplot2

### Preparing the Data

We first get the monthly sale totals for each product from 2013 onward.

```{r}
sales_report <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  mutate(Month = month(Date)) %>%
  group_by(Month, Product) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup() %>%
  select(Month, Product, Sales)
```

We then use the monthly sale totals data to compute the minimum, maximum and latest month sales.

```{r}
mins <- group_by(sales_report, Product) %>% 
  slice(which.min(Sales))
maxs <- group_by(sales_report, Product) %>% 
  slice(which.max(Sales))
ends <- group_by(sales_report, Product) %>% 
  filter(Month == max(Month))
```

We then compute the 25th and 75th quantiles.

```{r}
quarts <- sales_report %>%
  group_by(Product) %>%
  summarise(quart1 = quantile(Sales, 
                              0.25),
            quart2 = quantile(Sales, 
                              0.75)) %>%
  right_join(sales_report)
```

### Plotting the sparkline

```{r, warning=FALSE}
ggplot(sales_report, aes(x=Month, y=Sales)) + 
  facet_grid(Product ~ ., scales = "free_y") + 
  geom_ribbon(data = quarts, aes(ymin = quart1, max = quart2), 
              fill = 'grey90') +
  geom_line(size=0.3) +
  geom_point(data = mins, col = 'red') +
  geom_point(data = maxs, col = 'blue') +
  geom_text(data = mins, aes(label = Sales), vjust = -1, size = 3) +
  geom_text(data = maxs, aes(label = Sales), vjust = 2.5, size = 3) +
  geom_text(data = ends, aes(label = Sales), hjust = 0, nudge_x = 0.5, size = 2.5) +
  geom_text(data = ends, aes(label = Product), hjust = 0, nudge_x = 1.0, size = 2.5) +
  expand_limits(x = max(sales_report$Month) + 
                  (0.25 * (max(sales_report$Month) - min(sales_report$Month)))) +
  scale_x_continuous(breaks = seq(1, 12, 1)) +
  scale_y_continuous(expand = c(0.1, 0)) +
  theme_tufte(base_size = 3, base_family = "Helvetica") +
  theme(axis.title=element_blank(), axis.text.y = element_blank(), 
        axis.ticks = element_blank(), strip.text = element_blank())
```

## Static Information Dashboard Design: gt and gtExtras methods

In this section, we will create static information dashboards by using **gt** and **gtExtras** packages.

### Plotting a simple bullet chart

Here, we prepare a bullet chart report by using `gt()` and `gt_plt_bullet()` functions.

```{r}
product %>%
  gt::gt() %>%
  gt_plt_bullet(column = current, 
              target = target, 
              width = 60,
              palette = c("lightblue", 
                          "black")) %>%
  gt_theme_538()
```

## sparklines: gtExtras method

We first filter sales data for 2013 and calculate total monthly sales for each product.

```{r}
report <- coffeechain %>%
  mutate(Year = year(Date)) %>%
  filter(Year == "2013") %>%
  mutate (Month = month(Date, 
                        label = TRUE, 
                        abbr = TRUE)) %>%
  group_by(Product, Month) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup()
```

As **gtExtras** functions require us to pass data.frames with list columns, we first convert the report data.frame into list columns as below

```{r}
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop")
```

### Plotting Coffechain Sales report

```{r}
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop") %>%
   gt() %>%
   gt_plt_sparkline('Monthly Sales',
                    same_limit = FALSE)
```

### Adding statistics

We calculate some summary statistics as below

```{r}
report %>% 
  group_by(Product) %>% 
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            ) %>%
  gt() %>%
  fmt_number(columns = 4,
    decimals = 2)
```

### Combining the data.frame

We use the code below to add statistics to the table

```{r}
spark <- report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop")
```

```{r}
sales <- report %>% 
  group_by(Product) %>% 
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            )
```

```{r}
sales_data = left_join(sales, spark)
```

### Plot the updated data.table

```{r}
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales',
                   same_limit = FALSE)
```

## Combining bullet chart and sparklines

```{r}
bullet <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  group_by(`Product`) %>%
  summarise(`Target` = sum(`Budget Sales`),
            `Actual` = sum(`Sales`)) %>%
  ungroup() 
```

```{r}
sales_data = sales_data %>%
  left_join(bullet)
```

```{r}
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales') %>%
  gt_plt_bullet(column = Actual, 
                target = Target, 
                width = 28,
                palette = c("lightblue", 
                          "black")) %>%
  gt_theme_538()
```

### Interactive Information Dashboard Design: reactable and reactablefmtr methods

In this section, we will create interactive information dashboards by using [**reactable**](https://glin.github.io/reactable/index.html) and [**reactablefmtr**](https://kcuilla.github.io/reactablefmtr/index.html) packages.

In order to build interactive sparklines, we need to install [**dataui**](https://github.com/timelyportfolio/dataui) R package.

```{r}
remotes::install_github("timelyportfolio/dataui")
```
Load the package using a library call

```{r}
library(dataui)
```

### Plotting Interactive Sparklines

Similar to gtExtras, to plot an interactive sparklines by using **reactablefmtr** package, we need to prepare the list field.

```{r}
report <- report %>%
  group_by(Product) %>%
  summarize(`Monthly Sales` = list(Sales))
```

And then use [`react_sparkline()`](https://kcuilla.github.io/reactablefmtr/reference/react_sparkline.html) to plot the sparklines as shown below.

```{r}
reactable(
  report,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```
### Changing Pagesize

By default the pagesize is 10 i.e. the number of results returned per page. We add argument `defaultPageSize` to change the default setting.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```
### Adding points and labels

`highlight_points` argument is used to show the minimum and maximum values' points and `label` argument is used to label first and last values.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        labels = c("first", "last")
        )
    )
  )
)
```

### Adding reference lines

`statline` argument is used to show the mean line.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        statline = "mean"
        )
    )
  )
)
```

### Adding bandline

Instead of adding a reference line, a reference band can be added by using the `bandline` argument.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        line_width = 1,
        bandline = "innerquartiles",
        bandline_color = "green"
        )
    )
  )
)
```

### Changing from sparkline to sparkbar

Instead of displaying the values as sparklines, we can display them as sparkbars as shown below.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkbar(
        report,
        highlight_bars = highlight_bars(
          min = "red", max = "blue"),
        bandline = "innerquartiles",
        statline = "mean")
    )
  )
)
```

